---
title: "R Notebook"
output: pdf
---

#read environmental exposure data
```{r}
exp_dat<- read_csv("~/Documents/ECHO/ECHO_data/05_assays/Lab_Bio_Analysis_Results.csv") |>
  mutate(analysis_analyte_name = str_to_lower(analysis_analyte_name),
         log_lod = log2(analysis_llod + 1e-6))|>
  mutate(analysis_analyte_name = fct_recode(analysis_analyte_name,
                                            "bis(butoxyethyl) phosphate" = "bis(butoxyethyl) pohosphate",
                                            "bis(2-ethylhexyl) phosphate" = "bis(2-ethylhexyl) phoshate",
                                            "bis(1,3-dichloro-2-propyl) phosphate" = "bis(1,3-dichloro-2propyl) phosphate",
                                            "bis(1-chloro-2-propyl) phosphate" = "bis-(1-chloro-2propl) phosphate",
                                            "bis(2-chloroethyl) phosphate" = "bis-2chloroethyl phosphate")) 
```


#summarizing using filtered data
```{r}
exp_dat<- exp_dat_df |>
  filter(specimen_collection_trimester == 2)
```


#May 29
########
#######
#####
```{r}
# Step 1: Calculate summary statistics
summary_stats <- exp_dat %>%
  mutate(analysis_result = if_else(analysis_result < analysis_llod, analysis_llod / sqrt(2), analysis_result)) %>%
  group_by(analysis_analyte_name, analysis_llod) %>%
  summarise(
    log_lod_num = log2(first(analysis_llod)),  # Calculate log_lod as numeric
    std_dev_result_num = sd(log2(analysis_result), na.rm = TRUE), # Calculate std_dev_result as numeric
    half_std_dev_result = (sd(log2(analysis_result), na.rm = TRUE))/2,
    quart_std_dev_result = (sd(log2(analysis_result), na.rm = TRUE))/4,
    .groups = 'drop'  # Ungroup after summarising
  ) %>%
  mutate(
    lod_plus_sd_num = log_lod_num + std_dev_result_num,  # Calculate lod_plus_sd as numeric
    lod_minus_sd_num = log_lod_num - std_dev_result_num,  # Calculate lod_plus_sd as numeric
    lod_plus_half_sd = log_lod_num + half_std_dev_result,  
    lod_minus_half_sd = log_lod_num - half_std_dev_result, 
    lod_plus_quart_sd = log_lod_num + quart_std_dev_result,  
    lod_minus_quart_sd = log_lod_num - quart_std_dev_result
  )

# Step 2: Count observations within the specified range
count_within_range <- exp_dat %>%
  mutate(analysis_result = if_else(analysis_result < analysis_llod, analysis_llod / sqrt(2), analysis_result)) %>%
  left_join(summary_stats, by = c("analysis_analyte_name", "analysis_llod")) %>%
  group_by(analysis_analyte_name, analysis_llod) %>%
  summarise(
    count_lod_plus_sd = sum(log2(analysis_result) > log2(analysis_llod) & log2(analysis_result) < lod_plus_sd_num, na.rm = TRUE),
    count_lod_minus_sd = sum(log2(analysis_result) < log2(analysis_llod) & log2(analysis_result) > lod_minus_sd_num, na.rm = TRUE),
  
    count_lod_plus_half_sd = sum(log2(analysis_result) > log2(analysis_llod) & log2(analysis_result) < lod_plus_half_sd, na.rm = TRUE),
    count_lod_minus_half_sd = sum(log2(analysis_result) < log2(analysis_llod) & log2(analysis_result) > lod_minus_half_sd, na.rm = TRUE),
    
    count_lod_plus_quart_sd = sum(log2(analysis_result) > log2(analysis_llod) & log2(analysis_result) < lod_plus_quart_sd, na.rm = TRUE),
    count_lod_minus_quart_sd = sum(log2(analysis_result) < log2(analysis_llod) & log2(analysis_result) > lod_minus_quart_sd, na.rm = TRUE),
    
    total_obs = n(),
    obs_with_LOD_count = sum(!is.na(analysis_llod)),
    obs_with_LOD_percentage = sprintf("%.2f%%", obs_with_LOD_count / total_obs * 100),
    obs_bel_LOD_count = sum(!is.na(analysis_llod) & analysis_result <= analysis_llod, na.rm = TRUE),
    obs_bel_LOD_percentage = sprintf("%.2f%%", obs_bel_LOD_count / total_obs * 100),
    analyte_result_NA_count = sum(is.na(analysis_result)),
    analyte_result_NA_percentage = sprintf("%.2f%%", analyte_result_NA_count / total_obs * 100),
    min_value = ifelse(sum(!is.na(analysis_result)) > 0, sprintf("%.2f", min(analysis_result, na.rm = TRUE)), "NA"),
    percentile_10 = sprintf("%.2f", quantile(analysis_result, 0.10, na.rm = TRUE)),
    percentile_25 = sprintf("%.2f", quantile(analysis_result, 0.25, na.rm = TRUE)),
    percentile_50 = sprintf("%.2f", quantile(analysis_result, 0.50, na.rm = TRUE)),
    percentile_75 = sprintf("%.2f", quantile(analysis_result, 0.75, na.rm = TRUE)),
    percentile_90 = sprintf("%.2f", quantile(analysis_result, 0.90, na.rm = TRUE)),
    max_value = ifelse(sum(!is.na(analysis_result)) > 0, sprintf("%.2f", max(analysis_result, na.rm = TRUE)), "NA"),
    .groups = 'drop'
  )

# Step 3: Merge results and format output
sd_threshold <- summary_stats |>
  left_join(count_within_range, by = c("analysis_analyte_name", "analysis_llod"))  |>
  filter(!is.na(analysis_llod))|>
  mutate(across(2:6, ~ sprintf("%.2f", .)))

write_csv(sd_threshold, "~/Documents/ECHO/ECHO_exposure_summary_july09.csv")
```



#new summary format suggested by Aimin
#single imputation using LOD/sqrt-2
#using chemical biomarkers that has a single value as LOD
```{r}
summary_df <- exp_dat |>
  mutate(analysis_result = if_else(analysis_result < analysis_llod, analysis_llod / sqrt(2), analysis_result)) |>
  group_by(analysis_analyte_name) |>
  summarise(
    unique_llod_values = paste(
      unique(paste0(sprintf("%.2f", analysis_llod[!is.na(analysis_llod)]), " ", analysis_result_unit[!is.na(analysis_llod)])),
      collapse = ", "
    ),
    total_obs = n(),
    obs_with_LOD_count = sum(!is.na(analysis_llod)),
    obs_with_LOD_percentage = sprintf("%.2f%%", obs_with_LOD_count / total_obs * 100),
    obs_bel_LOD_count = sum(!is.na(analysis_llod) & analysis_result <= analysis_llod, na.rm = TRUE),
    obs_bel_LOD_percentage = sprintf("%.2f%%", obs_bel_LOD_count / total_obs * 100),
    analyte_result_NA_count = sum(is.na(analysis_result)),
    analyte_result_NA_percentage = sprintf("%.2f%%", analyte_result_NA_count / total_obs * 100),
    
    #log_half_sd = sprintf("%.2f", 0.5 * (sd(log2(analysis_result), na.rm = TRUE))),
    #count_less_than_log_half_sd = sum(log2(analysis_result) > log2(analysis_llod) & log2(analysis_result) < log_half_sd , na.rm = TRUE),
    
    min_value = ifelse(sum(!is.na(analysis_result)) > 0, sprintf("%.2f", min(analysis_result, na.rm = TRUE)), "NA"),
    percentile_10 = sprintf("%.2f", quantile(analysis_result, 0.10, na.rm = TRUE)),
    percentile_25 = sprintf("%.2f", quantile(analysis_result, 0.25, na.rm = TRUE)),
    percentile_50 = sprintf("%.2f", quantile(analysis_result, 0.50, na.rm = TRUE)),
    percentile_75 = sprintf("%.2f", quantile(analysis_result, 0.75, na.rm = TRUE)),
    percentile_90 = sprintf("%.2f", quantile(analysis_result, 0.90, na.rm = TRUE)),
    max_value = ifelse(sum(!is.na(analysis_result)) > 0, sprintf("%.2f", max(analysis_result, na.rm = TRUE)), "NA")
  )

write_csv(summary_df, "~/Documents/ECHO/ECHO_exposure_summary_may28_v2.csv")
```


#distribution plots - filtering data for 8 biomarkers
```{r}
exposure_subset <- exp_dat |>
  mutate(analysis_llod = as.numeric(sprintf("%.2f", analysis_llod))) |>
  filter(analysis_analyte_name %in% c("bis(2-ethylhexyl) phosphate",
                                      "bis(butoxyethyl) phosphate"
                                      )) |>
  filter(!is.na(analysis_result)) |>
  filter(!is.na(analysis_llod)) |>
  select(c(13,16,17,19)) |>
  mutate(analysis_result = if_else(analysis_result < analysis_llod, analysis_llod/sqrt(2), analysis_result)) 
  #mutate(log2_result = ifelse(analysis_result > analysis_llod, log2(analysis_result + 0.0000001), analysis_result))
  #mutate(log2_result = log2(analysis_result + 0.000001)) 
```

#left-truncated imputation - August 6
```{r}
set.seed(2024)

exposure_subset<- exp_dat |>
  select(c(1,2,4,5,6,7,8,9)) |>
  #filter(analysis_analyte_name == "bis(1-chloro-2-propyl) phosphate") |>
  mutate(analysis_result = if_else(
    analysis_result < analysis_llod,
    runif(n(), min = 1e-4, max = analysis_llod),
    analysis_result
  ),
  analysis_result = as.numeric(sprintf("%.4f", analysis_result))) |>
  rename(subject.id = x_participant_id)

labres<- exposure_subset |>
  mutate(lod.flag = if_else(analysis_result < analysis_llod, 1, 0),
         test.result = analysis_result,
         log2.test.result = log2(analysis_result),
         test.name = analysis_analyte_name,
         test.time = "one")

LODmeansd.all <- labres %>% 
  group_by(test.name, test.time) %>% 
  summarise(meanlog2 = mean(log2.test.result, na.rm= TRUE), sdlog2 = sd(log2.test.result, na.rm=TRUE)) %>% 
  mutate(test.name = (paste0(test.name, ".t", test.time)))

#repeat below process for each analyte

#BCPP
dat_bcpp <- labres |> filter(test.name == "bis(1-chloro-2-propyl) phosphate")
dat_imp_bcpp <- impute(dat_bcpp, LODmeansd.all, "bis(1-chloro-2-propyl) phosphate.tone", "result")

#BCEP
dat_bcep <- labres |> filter(test.name == "bis(2-chloroethyl) phosphate")
dat_imp_bcep <- impute(dat_bcep, LODmeansd.all, "bis(2-chloroethyl) phosphate.tone", "result")

#BDCPP
dat_bdcpp <- labres |> filter(test.name == "bis(1,3-dichloro-2-propyl) phosphate")
dat_imp_bdcpp <- impute(dat_bdcpp, LODmeansd.all, "bis(1,3-dichloro-2-propyl) phosphate.tone", "result")

#BEHP
dat_behp <- labres |> filter(test.name == "bis(2-ethylhexyl) phosphate")
dat_imp_behp <- impute(dat_behp, LODmeansd.all, "bis(2-ethylhexyl) phosphate.tone", "result")

#BMPP
dat_bmpp <- labres |> filter(test.name == "bis(2-methylphenyl) phosphate")
dat_imp_bmpp <- impute(dat_bmpp, LODmeansd.all, "bis(2-methylphenyl) phosphate.tone", "result")

#BBEP
dat_bbep <- labres |> filter(test.name == "bis(butoxyethyl) phosphate")
dat_imp_bbep <- impute(dat_bbep, LODmeansd.all, "bis(butoxyethyl) phosphate.tone", "result")

#DEP
dat_dep <- labres |> filter(test.name == "diethylphosphate")
dat_imp_dep <- impute(dat_dep, LODmeansd.all, "diethylphosphate.tone", "result")

#DPP
dat_dpp <- labres |> filter(test.name == "dipropyl phosphate")
dat_imp_dpp <- impute(dat_dpp, LODmeansd.all, "dipropyl phosphate.tone", "result")
```

#join imputed datasets
```{r}
dat_imp_bcpp <- dat_imp_bcpp |> mutate(analyte = "bis(1-chloro-2-propyl) phosphate")
dat_imp_bcep <- dat_imp_bcep |> mutate(analyte = "bis(2-chloroethyl) phosphate")
dat_imp_bdcpp <- dat_imp_bdcpp |> mutate(analyte = "bis(1,3-dichloro-2-propyl) phosphate")
dat_imp_behp <- dat_imp_behp |> mutate(analyte = "bis(2-ethylhexyl) phosphate")
dat_imp_bmpp <- dat_imp_bmpp |> mutate(analyte = "bis(2-methylphenyl) phosphate")
dat_imp_bbep <- dat_imp_bbep |> mutate(analyte = "bis(butoxyethyl) phosphate")
dat_imp_dep <- dat_imp_dep |> mutate(analyte = "diethylphosphate")
dat_imp_dpp <- dat_imp_dpp |> mutate(analyte = "dipropyl phosphate")

imputed_dat<- bind_rows(dat_imp_bcpp, dat_imp_bcep, dat_imp_bdcpp,
                        dat_imp_behp, dat_imp_bmpp, dat_imp_bbep,
                        dat_imp_dep, dat_imp_dpp)
write_csv(imputed_dat, "~/Documents/ECHO/imputed_left_trunc_aug06.csv")
```

#prep for viz
```{r}
imputed_dat<- read_csv("~/Documents/ECHO/imputed_left_trunc_aug06.csv")

viz_dat<- imputed_dat |>
  mutate(log2_result= log2(result.res)) |>
  filter(analyte == "bis(1,3-dichloro-2-propyl) phosphate")


line_data <- tibble(
  xintercept = c(-4.069298652,	-7.218413728,	-4.856577421,	-6.431134959,	-5.250216805,	-6.037495574,	-5.6439),
  color = c("black", "black", "red", "red", "purple", "purple", "orange"),
  linetype = c("dashed", "dashed", "dashed", "dashed", "dashed", "dashed", "solid")  # Added linetype column
)

# Plotting the density plot with histogram faceted by analysis_analyte_name
viz_dat |>
  ggplot(aes(x = log2_result)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 0.25, fill = "skyblue", color = "skyblue", alpha = 0.5) +
  geom_density(color = "royalblue", linewidth=0.9) +
  labs(x = "Biomarker concentration on log2 scale", y = "Density",
       caption = " ") +
  ggtitle("") +
  scale_x_continuous(breaks = seq(-16, 6, by = 2), labels = seq(-16, 6, by = 2)) +
  theme_minimal() +
  geom_vline(data = line_data, aes(xintercept = xintercept, color = color, linetype = linetype), linewidth = 1) +
  scale_color_manual(values = c("black", "red", "purple", "orange")) +
  scale_linetype_manual(values = c("dashed", "solid")) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.caption = element_text(hjust = 0),  # Left-align the caption
    legend.position = "none"
  )

ggsave("~/Documents/ECHO/exposure_viz_aug_06.tiff",
       width=10, height= 10, dpi=300)
```

# Calculate the standard deviation of result.flg grouped by analyte
```{r}
std_dev_result <- imputed_dat |>
  group_by(analyte) |>
  summarise(std_dev_result.flg = 0.5*sd(log2(result.flag), na.rm = TRUE))

imputed_dat %>%
  filter(analyte == "bis(2-ethylhexyl) phosphate") |>
  filter(log2(result.flag) > -5.64 & log2(result.flag) < -4.43) %>%
  count()

imputed_dat %>%
  filter(analyte == "bis(2-ethylhexyl) phosphate") |>
  filter(log2(result.flag) < -5.64 & log2(result.flag) > -6.85) %>%
  count()
```




```{r}
# Define x-axis intercepts for each facet panel
x_intercepts <- data.frame(
  analyte = c("bis(2-ethylhexyl) phosphate", "bis(butoxyethyl) phosphate"),
  x_intercept = c(-5.643856, -5.643856)
)

# Plotting the density plot with histogram faceted by analysis_analyte_name
p <- viz_dat |>
  ggplot(aes(x = log2_result)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 0.1, fill = "skyblue", color = "skyblue", alpha = 0.5) +
  geom_density(color = "royalblue", linewidth=1) +
  labs(x = "Biomarker concentration on log2 scale", y = "Density",
       caption = " ") +
  ggtitle("") +
  #scale_y_continuous(trans = "exp")+
  #annotation_logticks(sides="lr")+
  scale_x_continuous(breaks = seq(-12, 5, by = 3), labels = seq(-12, 5, by = 3))+
  facet_wrap(~ analyte) +
  theme_minimal()+
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.caption = element_text(hjust = 0)  # Left-align the caption
  )

# Add vertical lines for each facet panel
p + geom_vline(data = x_intercepts, aes(xintercept = x_intercept), linetype = "dashed", color = "red", linewidth = 0.7, alpha = 0.8)

ggsave("~/Documents/ECHO/exposure_viz.tiff",
       width=10, height= 8, dpi=300)
```

```{r}
x <- exp_dat |>
  mutate(LOD_binary = ifelse(analysis_result > analysis_llod, 1, 0)) |>
  mutate(analysis_analyte_name = fct_recode(analysis_analyte_name,
                                            "bis(butoxyethyl) phosphate" = "bis(butoxyethyl) pohosphate",
                                            "bis(2-ethylhexyl) phosphate" = "bis(2-ethylhexyl) phoshate")) |>
  filter(analysis_analyte_name %in% c("bis(1-chloro-2-propyl) phosphate")) |>
  filter(!is.na(analysis_result)) |>
  filter(!is.na(analysis_llod)) |>
  select(c(10,13,16,17,19))
```


#summary with lef-truncated imputation - Aug 06
```{r}
dat<- read_csv("~/Documents/ECHO/imputed_left_trunc_aug06.csv") |>
  mutate(lod = case_when(
    analyte == "diethylphosphate" ~ 0.03,
    analyte == "bis(1,3-dichloro-2-propyl) phosphate" ~ 0.02,
    analyte == "bis(butoxyethyl) phosphate" ~ 0.02,
    analyte == "bis(2-chloroethyl) phosphate" ~ 0.02,
    analyte == "bis(1-chloro-2-propyl) phosphate" ~ 0.02,
    analyte == "bis(2-methylphenyl) phosphate" ~ 0.01,
    analyte == "bis(2-ethylhexyl) phosphate" ~ 0.02,
    analyte == "dipropyl phosphate" ~ 0.03,
    TRUE ~ NA_real_  # If analyte does not match any of the above, assign NA
  ))
```

#summary table
```{r}
# Step 1: Calculate summary statistics
summary_stats <- dat %>%
  group_by(analyte) %>%
  summarise(
    log_lod_num = log2(first(lod)),  # Calculate log_lod as numeric
    std_dev_result_num = sd(log2(result.res), na.rm = TRUE), # Calculate std_dev_result as numeric
    half_std_dev_result = (sd(log2(result.res), na.rm = TRUE))/2,
    quart_std_dev_result = (sd(log2(result.res), na.rm = TRUE))/4,
    eigth_std_dev_result = (sd(log2(result.res), na.rm = TRUE))/8,
    .groups = 'drop'  # Ungroup after summarising
  ) %>%
  mutate(
    lod_plus_sd_num = log_lod_num + std_dev_result_num,  # Calculate lod_plus_sd as numeric
    lod_minus_sd_num = log_lod_num - std_dev_result_num,  # Calculate lod_plus_sd as numeric
    lod_plus_half_sd = log_lod_num + half_std_dev_result,  
    lod_minus_half_sd = log_lod_num - half_std_dev_result, 
    lod_plus_quart_sd = log_lod_num + quart_std_dev_result,  
    lod_minus_quart_sd = log_lod_num - quart_std_dev_result,
    lod_plus_eigth_sd = log_lod_num + eigth_std_dev_result,  
    lod_minus_eigth_sd = log_lod_num - eigth_std_dev_result
  )

# Step 2: Count observations within the specified range
count_within_range <- dat %>%
  left_join(summary_stats, by = c("analyte")) %>%
  group_by(analyte) %>%
  summarise(
    count_lod_plus_sd = sum(log2(result.res) > log2(lod) & log2(result.res) < lod_plus_sd_num, na.rm = TRUE),
    count_lod_minus_sd = sum(log2(result.res) < log2(lod) & log2(result.res) > lod_minus_sd_num, na.rm = TRUE),
  
    count_lod_plus_half_sd = sum(log2(result.res) > log2(lod) & log2(result.res) < lod_plus_half_sd, na.rm = TRUE),
    count_lod_minus_half_sd = sum(log2(result.res) < log2(lod) & log2(result.res) > lod_minus_half_sd, na.rm = TRUE),
    
    count_lod_plus_quart_sd = sum(log2(result.res) > log2(lod) & log2(result.res) < lod_plus_quart_sd, na.rm = TRUE),
    count_lod_minus_quart_sd = sum(log2(result.res) < log2(lod) & log2(result.res) > lod_minus_quart_sd, na.rm = TRUE),
    
    count_lod_plus_eigth_sd = sum(log2(result.res) > log2(lod) & log2(result.res) < lod_plus_eigth_sd, na.rm = TRUE),
    count_lod_minus_eigth_sd = sum(log2(result.res) < log2(lod) & log2(result.res) > lod_minus_eigth_sd, na.rm = TRUE),
    
    total_obs = n(),
    obs_with_LOD_count = sum(!is.na(lod)),
    obs_with_LOD_percentage = sprintf("%.4f%%", obs_with_LOD_count / total_obs * 100),
    obs_bel_LOD_count = sum(!is.na(lod) & result.res <= lod, na.rm = TRUE),
    obs_bel_LOD_percentage = sprintf("%.4f%%", obs_bel_LOD_count / total_obs * 100),
    analyte_result_NA_count = sum(is.na(result.res)),
    analyte_result_NA_percentage = sprintf("%.4f%%", analyte_result_NA_count / total_obs * 100),
    min_value = ifelse(sum(!is.na(result.res)) > 0, sprintf("%.4f", min(result.res, na.rm = TRUE)), "NA"),
    percentile_10 = sprintf("%.4f", quantile(result.res, 0.10, na.rm = TRUE)),
    percentile_25 = sprintf("%.4f", quantile(result.res, 0.25, na.rm = TRUE)),
    percentile_50 = sprintf("%.4f", quantile(result.res, 0.50, na.rm = TRUE)),
    percentile_75 = sprintf("%.4f", quantile(result.res, 0.75, na.rm = TRUE)),
    percentile_90 = sprintf("%.4f", quantile(result.res, 0.90, na.rm = TRUE)),
    max_value = ifelse(sum(!is.na(result.res)) > 0, sprintf("%.4f", max(result.res, na.rm = TRUE)), "NA"),
    .groups = 'drop'
  )

# Step 3: Merge results and format output
sd_threshold <- summary_stats |>
  left_join(count_within_range, by = c("analyte"))  |>
  mutate(across(2:6, ~ sprintf("%.4f", .)))

write_csv(sd_threshold, "~/Documents/ECHO/summary_aug_06.csv")
```

#exposure differences by demographic groups
```{r}
imputed_dat<- read_csv("~/Documents/ECHO/imputed_left_trunc_aug06.csv") |>
  clean_names() |>
  mutate(truncated_id = substr(subject_id, 1, 6))

#age variable from Der file
dem_age <- read_csv("~/Documents/ECHO/ECHO_data/demog/Der_Dem_DemChild.csv") |>
  clean_names() |>
  select(c(1,2,6,31))


#ESS file
dem_b <- read_csv("~/Documents/ECHO/ECHO_data/demog/Ess_Dem_Dem_B.csv") |>
  select(c(4,5,6,10,12,13:28,45)) |>
  clean_names() |>
  mutate(
    x_form_dt = mdy(x_form_dt), 
    xdem_b_a1 = mdy(xdem_b_a1), 
    duration_years = sprintf("%.1f", as.numeric(difftime(xdem_b_a1, x_form_dt, units = "days")) / 365.25)
  ) |>
  mutate(
    dem_b_a5 = recode(dem_b_a5, `1` = "Yes", `2` = "No",
                        `-6` = "Not_applicable", `-7` = "Pref_not_answ", 
                        `-8` = "dont_know", `-9` = "missing"),
    dem_b_a2 = recode(dem_b_a2,
                       `1` = "No schooling/never attended",
                       `2` = "8th grade or less",
                       `3` = "Some high school, no degree",
                       `4` = "High school degree",
                       `5` = "GED or equivalent",
                       `6` = "Some college, no degree",
                       `7` = "Associate's degree (AA, AS)",
                       `8` = "Bachelor's degree (BA, BS)",
                       `9` = "Master's degree (MA, MS, MEd, MSW, MBA, MPH)",
                       `10` = "Professorial or Doctorate Degree (PhD, EdD, MD, JD)",
                       `-6` = "Not applicable",
                       `-7` = "Prefer not to answer",
                       `-8` = "Don't know",
                       `-9` = "Missing"),
    dem_b_a3 = recode(dem_b_a3,
                      `1` = "Yes",
                      `2` = "No",
                      `-6` = "Not applicable",
                      `-7` = "Prefer not to answer",
                      `-8` = "Don't know",
                      `-9` = "Missing"),
    dem_b_a4_1 = recode(dem_b_a4_1, `0` = "non_white", `1` = "white"),
    dem_b_a4_2 = recode(dem_b_a4_2, `0` = "non_afr_american", `1` = "afr_american"),
    dem_b_a4_3 = recode(dem_b_a4_3, `0` = "non_ameri_indi_alask", `1` = "ameri_indi_alask"),
    dem_b_a4_4 = recode(dem_b_a4_4, `0` = "non_asia_ameri", `1` = "asia_ameri"),
    dem_b_a4_5 = recode(dem_b_a4_5, `0` = "non_othr_asian", `1` = "othr_asian"),
    dem_b_a4_6 = recode(dem_b_a4_6, `0` = "non_NHOPI", `1` = "NHOPI"),
    race = case_when(
    # Check each column starting with 'dem_b_a4' and assign the value if it doesn't start with 'non_'
    !is.na(dem_b_a4_1) & !str_starts(dem_b_a4_1, "non_") ~ dem_b_a4_1,
    !is.na(dem_b_a4_2) & !str_starts(dem_b_a4_2, "non_") ~ dem_b_a4_2,
    !is.na(dem_b_a4_3) & !str_starts(dem_b_a4_3, "non_") ~ dem_b_a4_3,
    !is.na(dem_b_a4_4) & !str_starts(dem_b_a4_4, "non_") ~ dem_b_a4_4,
    !is.na(dem_b_a4_5) & !str_starts(dem_b_a4_5, "non_") ~ dem_b_a4_5,
    !is.na(dem_b_a4_6) & !str_starts(dem_b_a4_6, "non_") ~ dem_b_a4_6,
    TRUE ~ NA_character_  # If none match, assign NA
  )
  ) |>
  select(c(1,2,5,6,22,24)) |>
  rename(educa = dem_b_a2,
         hisp_ori = dem_b_a3,
         foreg_brn = dem_b_a5) 
#not considered Age (in years) at time of form completion since it has values form -1 to 21

#removing duplicates
dem_b_clean <- dem_b |>
    mutate(truncated_id = substr(x_participant_id, 1, 6)) |>
    rowwise() |>
    mutate(completeness_score = sum(!is.na(c(educa, hisp_ori, foreg_brn, race)))) |>
    ungroup() |>
    group_by(truncated_id) |>
    arrange(desc(completeness_score), visit_name) |>
    slice_head(n = 1) |>
    select(-truncated_id, -completeness_score)

#join demographics and exposure data
demog_exp<- left_join(imputed_dat, dem_b_clean,
                      by="truncated_id")

#export data - to test demographic differences - August 23, 2024
write_csv(demog_exp, "~/Documents/ECHO/ECHO_data/derived_df/demog_exp.csv")

```

### For AGE VARIABLE
```{r}
#Der file
dem_b <- read_csv("~/Documents/ECHO/ECHO_data/demog/Der_Dem_DemChild.csv") |>
  clean_names() |>
  select(c(1,2,31)) |>
  mutate(demchild_mat_age = ifelse(demchild_mat_age == -9, NA, demchild_mat_age),
         truncated_id = substr(x_preg_id, 1, 6)) |>
  select(c(4,3)) |>
  group_by(truncated_id) |> 
  summarise(demchild_mat_age = as.numeric(sprintf("%.1f", mean(demchild_mat_age, na.rm = TRUE)))) |> 
  ungroup()
  
  

ggplot(dem_b, aes(x = demchild_mat_age)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 5, fill = "royalblue", color = "black", alpha = 0.5) +
  geom_density(color = "blue", linewidth = 1) +
  labs(title = " ", x = "Age", y = "Density") +
  scale_x_continuous(breaks = seq(10,70, by = 10)) +
  theme_minimal()

#join this with demog_exp.csv
df<- left_join(demog_exp, dem_b, by= "truncated_id")
write_csv(df, "~/Documents/ECHO/ECHO_data/derived_df/demog_exp_v_sep_7_2024.csv")
```

#testing for group differences - Sep 7, 2024
#formatting steps
```{r}
demog_exp<- read_csv("~/Documents/ECHO/ECHO_data/derived_df/demog_exp_v_sep_7_2024.csv") |>
  mutate(
    race = recode(race, "white" = "white", .default = "non_white") |> 
      fct_na_value_to_level("non_white"),
    hisp_ori = recode(hisp_ori, "Yes" = "yes", .default = "no") |> 
      fct_na_value_to_level("no"),
    foreg_brn = recode(foreg_brn, "Yes" = "yes", .default = "no") |> 
      fct_na_value_to_level("no")
  ) |>
  mutate(
    educa = case_when(
      educa %in% c(
        "Bachelor's degree (BA, BS)", 
        "Master's degree (MA, MS, MEd, MSW, MBA, MPH)",
        "Professorial or Doctorate Degree (PhD, EdD, MD, JD)",
        "High school degree","Associate's degree (AA, AS)",
        "Some high school, no degree")
      ~ "high_school_or_abv",
      educa %in% c("8th grade or less", 
        "Don't know","GED or equivalent", 
        "Prefer not to answer",
        "Some college, no degree") ~ "less_than_high_school", TRUE 
      ~ "less_than_high_school")) |>
  mutate(truncated_id = substr(subject_id, 1, 9))


#join outcome here
demog_exp_out<- left_join(demog_exp, dt_out, by="truncated_id") 

#join mom outcomes
final_df<- left_join(demog_exp_out, mom_dat_imp, by="truncated_id")
```


#data version saved on Nov 07
```{r}
write_csv(final_df, "~/Documents/ECHO/ECHO_data/derived_df/demog_exp_out_nov_22.csv")
demog_exp_out<- read_csv("~/Documents/ECHO/ECHO_data/derived_df/demog_exp_out_nov_22.csv")
```

#filter strategy update September 20
```{r}
#bis(1,3-dichloro-2-propyl) phosphate; LOD 0.02; 13% below LOD
#bis(2-chloroethyl) phosphate; LOD 0.02; 32% below LOD
#bis(1-chloro-2-propyl) phosphate; LOD 0.02; 43% below LOD
#bis(2-ethylhexyl) phosphate; LOD 0.02; 73% below LOD
#bis(butoxyethyl) phosphate; LOD 0.02; 33.2% below LOD
#bis(2-methylphenyl) phosphate; LOD 0.01; 63.7% below LOD

#diethylphosphate; LOD 0.03; 33.2% below LOD ##
#dipropyl_phosphate; LOD 0.03; 33.2% below LOD ##

chem<- "dipropyl phosphate"
lod_value<- 0.03

bcpp_filtered<- demog_exp_out |>
  filter(analyte %in% c(chem)) |>
  filter(result_res >= lod_value-lod_value/3 & 
           result_res <= lod_value+lod_value/3) |> #divide by here
  mutate(category = case_when(
    result_res < lod_value ~ "below_lod",
    result_res >= lod_value ~ "above_lod"
  ),
  demchild_mat_age = if_else(is.na(demchild_mat_age), median(demchild_mat_age, na.rm = TRUE), demchild_mat_age),
    birth_ga = if_else(is.na(birth_ga), median(birth_ga, na.rm = TRUE), birth_ga),
    birth_bw = if_else(is.na(birth_bw), median(birth_bw, na.rm = TRUE), birth_bw),
  birth_bw_c = if_else(is.na(birth_bw_c), "normal", birth_bw_c),
    birth_ga_c = if_else(is.na(birth_ga_c), "not_PTB", birth_ga_c),
    sga = if_else(is.na(sga), "not_SGA", sga))
```

#sd based filtering
```{r}

bcpp_filtered<- demog_exp_out |>
  filter(analyte %in% c(chem)) |>
  filter(result_res >= 2^(-5.717800746) & #minus_half_sd_value
           result_res <= 2^(-4.399986632)) |> #plus_half_sd_value
  mutate(category = case_when(
    result_res < lod_value ~ "below_lod",
    result_res >= lod_value ~ "above_lod"
  ),
  demchild_mat_age = if_else(is.na(demchild_mat_age), median(demchild_mat_age, na.rm = TRUE), demchild_mat_age),
    birth_ga = if_else(is.na(birth_ga), median(birth_ga, na.rm = TRUE), birth_ga),
    birth_bw = if_else(is.na(birth_bw), median(birth_bw, na.rm = TRUE), birth_bw),
  birth_bw_c = if_else(is.na(birth_bw_c), "normal", birth_bw_c),
    birth_ga_c = if_else(is.na(birth_ga_c), "not_PTB", birth_ga_c),
    sga = if_else(is.na(sga), "not_SGA", sga))
```



```{r}
# Call the function with your dataframe and variables
results_df <- generate_tables(df=bcpp_filtered, 
                              variables= c("race", "hisp_ori", 
                                           "educa", "foreg_brn",
                                           "birth_bw_c", "birth_ga_c", "sga",
                                           "demchild_mat_age",
                                           "birth_bw", "birth_ga",
                                           "cond_gdm", "cond_ghtn", "cond_pre"), 
                              category="category")

for (var in names(results_df)) {
  cat(paste(results_df[[var]], collapse = "\n"))
  cat("\n")
}
```

#mean difference of other exposures
```{r}
mean_diff_df <- demog_exp_out |> 
  semi_join(bcpp_filtered, by = "subject_id") |>
  group_by(subject_id) |> 
  mutate(category = if_else(any(analyte == chem & result_res < lod_value), 
                            "below_lod", 
                            "above_lod")) |> 
  ungroup() |>
  mutate(result_res = log2(result_res))

results<- analyte_summary(mean_diff_df, analyte, category, result_res)

for (analyte in names(results)) {
  cat(paste(results[[analyte]], collapse = "\n"))
  cat("\n")
}
```

#correlation heatmap
```{r}
demog_exp<- read_csv("~/Documents/ECHO/ECHO_data/derived_df/demog_exp_v_sep_7_2024.csv") |>
  select(c(1,3,4)) |>
  mutate(result_res=as.numeric(log2(result_res)))

# Reshape data from long to wide format
wide_data <- demog_exp %>%
  group_by(subject_id, analyte) %>%
  summarise(result_res = mean(result_res, na.rm = TRUE), .groups = 'drop') |>
  pivot_wider(names_from = analyte, values_from = result_res) |>
  select(-c(1)) |>
  mutate(across(everything(), as.numeric))

cormat <- round(x=cor(wide_data, method = "spearman", use = "complete.obs"), digits = 2) |>
  melt() |>
  clean_names()
```

```{r}
ggplot(cormat, aes(x = var2, y = var1, fill = value, label = value)) + #, 
    geom_tile(color = "white") +
    geom_text(color = "black", size = 5, vjust = 1) +
    scale_fill_gradient2(low = "red", high = "royalblue", mid = "white",
                         midpoint = 0,
                         limit = c(-1, 1), space = "Lab",
                         name = "Spearman Correlation coefficient") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 11, hjust = 1),
          axis.text.y = element_text(angle = 0, vjust = 0.5, size = 11, hjust = 1),
          axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          panel.grid.major = element_blank(),
          panel.border = element_blank(),
          panel.background = element_blank(),
          axis.ticks = element_blank(),
          legend.position = "bottom", legend.box = "horizontal") +
    coord_fixed()
```
======================================================
race
------------------------------------------------------
                 above_lod            below_lod
------------------------------------------------------
non_white  |   1 (2.0%, 33.3%)  |   48 (98.0%, 49.0%)
white      |   2 (3.8%, 66.7%)  |   50 (96.2%, 51.0%)
------------------------------------------------------
race: chi_square statistic: 2.014177e-31 (p-value: 1)
race: fisher p_value: 1
========================================================
