---
title: "R Notebook"
output: pdf
---

#read environmental exposure data
```{r}
exp_dat<- read_csv("~/Documents/ECHO/ECHO_data/05_assays/Lab_Bio_Analysis_Results.csv") |>
  mutate(analysis_analyte_name = str_to_lower(analysis_analyte_name),
         log_lod = log2(analysis_llod + 1e-6))|>
  mutate(analysis_analyte_name = fct_recode(analysis_analyte_name,
                                            "bis(butoxyethyl) phosphate" = "bis(butoxyethyl) pohosphate",
                                            "bis(2-ethylhexyl) phosphate" = "bis(2-ethylhexyl) phoshate",
                                            "bis(1,3-dichloro-2-propyl) phosphate" = "bis(1,3-dichloro-2propyl) phosphate",
                                            "bis(1-chloro-2-propyl) phosphate" = "bis-(1-chloro-2propl) phosphate",
                                            "bis(2-chloroethyl) phosphate" = "bis-2chloroethyl phosphate")) 

|>
  filter(grepl("phosphate", analysis_analyte_name, ignore.case = TRUE))
  
```

#summary of exposure data
```{r}
summary_df <- exp_dat |>
  group_by(analysis_analyte_name) |>
  summarise(
    total_obs = n(),
    obs_with_LOD = paste(
      sum(!is.na(analysis_llod)),
      sprintf("(%.2f%%)", sum(!is.na(analysis_llod)) / total_obs * 100)
    ),
    count_without_LOD = sum(is.na(analysis_llod)),
    pct_without_LOD = sprintf("(%.2f%%)", sum(is.na(analysis_llod)) / total_obs * 100),
    
    obs_bel_LOD = paste(
      sum(!is.na(analysis_llod) & analysis_result <= analysis_llod, na.rm = TRUE),
      sprintf("(%.2f%%)", sum(!is.na(analysis_llod) & analysis_result <= analysis_llod, na.rm = TRUE) / total_obs * 100)
    ),
    unique_llod_values = paste(
      unique(paste0(sprintf("%.2f", analysis_llod[!is.na(analysis_llod)]), " ", analysis_result_unit[!is.na(analysis_llod)])),
      collapse = ", "
    ),
    analyte_result_NA = paste(
      sum(is.na(analysis_result)),
      sprintf("(%.2f%%)", sum(is.na(analysis_result)) / total_obs * 100)
    ),
    min_value = ifelse(sum(!is.na(analysis_result[analysis_result > analysis_llod])) > 0, sprintf("%.2f", min(analysis_result[analysis_result > analysis_llod], na.rm = TRUE)), "NA"),
    percentile_10 = sprintf("%.2f", quantile(analysis_result[analysis_result > analysis_llod], 0.10, na.rm = TRUE)),
    percentile_25 = sprintf("%.2f", quantile(analysis_result[analysis_result > analysis_llod], 0.25, na.rm = TRUE)),
    percentile_50 = sprintf("%.2f", quantile(analysis_result[analysis_result > analysis_llod], 0.50, na.rm = TRUE)),
    percentile_75 = sprintf("%.2f", quantile(analysis_result[analysis_result > analysis_llod], 0.75, na.rm = TRUE)),
    percentile_90 = sprintf("%.2f", quantile(analysis_result[analysis_result > analysis_llod], 0.90, na.rm = TRUE)),
    max_value = ifelse(sum(!is.na(analysis_result[analysis_result > analysis_llod])) > 0, sprintf("%.2f", max(analysis_result[analysis_result > analysis_llod], na.rm = TRUE)), "NA")
  )

write_csv(summary_df, "~/Documents/ECHO/ECHO_exposure_summary.csv")
```

#new summary format
```{r}
summary_df <- exp_dat |>
  group_by(analysis_analyte_name) |>
  summarise(
    total_obs = n(),
    obs_with_LOD_count = sum(!is.na(analysis_llod)),
    obs_with_LOD_percentage = sprintf("%.2f%%", obs_with_LOD_count / total_obs * 100),
    obs_without_LOD_count = sum(is.na(analysis_llod)),
    obs_without_LOD_percentage = sprintf("%.2f%%", obs_without_LOD_count / total_obs * 100),
    obs_bel_LOD_count = sum(!is.na(analysis_llod) & analysis_result <= analysis_llod, na.rm = TRUE),
    obs_bel_LOD_percentage = sprintf("%.2f%%", obs_bel_LOD_count / total_obs * 100),
    analyte_result_NA_count = sum(is.na(analysis_result)),
    analyte_result_NA_percentage = sprintf("%.2f%%", analyte_result_NA_count / total_obs * 100),
    unique_llod_values = paste(
      unique(paste0(sprintf("%.2f", analysis_llod[!is.na(analysis_llod)]), " ", analysis_result_unit[!is.na(analysis_llod)])),
      collapse = ", "
    ),
    log_sd = sprintf("%.2f", 0.5*(sd(analysis_result[analysis_result > analysis_llod], na.rm = TRUE))),
    log_half_sd = sprintf("%.2f", 0.5*(sd(analysis_result[analysis_result > analysis_llod], na.rm = TRUE))),
    log_quat_sd = sprintf("%.2f", 0.25*(sd(analysis_result[analysis_result > analysis_llod], na.rm = TRUE))),
    min_value = ifelse(sum(!is.na(analysis_result[analysis_result > analysis_llod])) > 0, sprintf("%.2f", min(analysis_result[analysis_result > analysis_llod], na.rm = TRUE)), "NA"),
    percentile_10 = sprintf("%.2f", quantile(analysis_result[analysis_result > analysis_llod], 0.10, na.rm = TRUE)),
    percentile_25 = sprintf("%.2f", quantile(analysis_result[analysis_result > analysis_llod], 0.25, na.rm = TRUE)),
    percentile_50 = sprintf("%.2f", quantile(analysis_result[analysis_result > analysis_llod], 0.50, na.rm = TRUE)),
    percentile_75 = sprintf("%.2f", quantile(analysis_result[analysis_result > analysis_llod], 0.75, na.rm = TRUE)),
    percentile_90 = sprintf("%.2f", quantile(analysis_result[analysis_result > analysis_llod], 0.90, na.rm = TRUE)),
    max_value = ifelse(sum(!is.na(analysis_result[analysis_result > analysis_llod])) > 0, sprintf("%.2f", max(analysis_result[analysis_result > analysis_llod], na.rm = TRUE)), "NA")
  )

write_csv(summary_df, "~/Documents/ECHO/ECHO_exposure_summary.csv")
```

#new summary format suggested by Aimin
#single imputation using LOD/sqrt-2
#using chemical biomarkers that has a single value as LOD
```{r}
threshold_df <- exp_dat |>
  mutate(analysis_result = if_else(analysis_result < analysis_llod, analysis_llod / sqrt(2), analysis_result)) |>
  group_by(analysis_analyte_name, analysis_llod) 


summary_df <- exp_dat |>
  mutate(analysis_result = if_else(analysis_result < analysis_llod, analysis_llod / sqrt(2), analysis_result)) |>
  group_by(analysis_analyte_name) |>
  summarise(
    unique_llod_values = paste(
      unique(paste0(sprintf("%.2f", analysis_llod[!is.na(analysis_llod)]), " ", analysis_result_unit[!is.na(analysis_llod)])),
      collapse = ", "
    ),
    total_obs = n(),
    obs_with_LOD_count = sum(!is.na(analysis_llod)),
    obs_with_LOD_percentage = sprintf("%.2f%%", obs_with_LOD_count / total_obs * 100),
    obs_bel_LOD_count = sum(!is.na(analysis_llod) & analysis_result <= analysis_llod, na.rm = TRUE),
    obs_bel_LOD_percentage = sprintf("%.2f%%", obs_bel_LOD_count / total_obs * 100),
    analyte_result_NA_count = sum(is.na(analysis_result)),
    analyte_result_NA_percentage = sprintf("%.2f%%", analyte_result_NA_count / total_obs * 100),
    
    #log_half_sd = sprintf("%.2f", 0.5 * (sd(log2(analysis_result), na.rm = TRUE))),
    #count_less_than_log_half_sd = sum(log2(analysis_result) > log2(analysis_llod) & log2(analysis_result) < log_half_sd , na.rm = TRUE),
    
    min_value = ifelse(sum(!is.na(analysis_result)) > 0, sprintf("%.2f", min(analysis_result, na.rm = TRUE)), "NA"),
    percentile_10 = sprintf("%.2f", quantile(analysis_result, 0.10, na.rm = TRUE)),
    percentile_25 = sprintf("%.2f", quantile(analysis_result, 0.25, na.rm = TRUE)),
    percentile_50 = sprintf("%.2f", quantile(analysis_result, 0.50, na.rm = TRUE)),
    percentile_75 = sprintf("%.2f", quantile(analysis_result, 0.75, na.rm = TRUE)),
    percentile_90 = sprintf("%.2f", quantile(analysis_result, 0.90, na.rm = TRUE)),
    max_value = ifelse(sum(!is.na(analysis_result)) > 0, sprintf("%.2f", max(analysis_result, na.rm = TRUE)), "NA")
  )

write_csv(summary_df, "~/Documents/ECHO/ECHO_exposure_summary_may28_v2.csv")
```


#distribution plots - filtering data for 8 biomarkers
```{r}
exposure_subset <- exp_dat |>
  mutate(analysis_llod = as.numeric(sprintf("%.2f", analysis_llod))) |>
  filter(analysis_analyte_name %in% c("bis(2-ethylhexyl) phosphate",
                                      "bis(butoxyethyl) phosphate"
                                      )) |>
  filter(!is.na(analysis_result)) |>
  filter(!is.na(analysis_llod)) |>
  select(c(13,16,17,19)) |>
  mutate(analysis_result = if_else(analysis_result < analysis_llod, analysis_llod/sqrt(2), analysis_result)) 
  #mutate(log2_result = ifelse(analysis_result > analysis_llod, log2(analysis_result + 0.0000001), analysis_result))
  #mutate(log2_result = log2(analysis_result + 0.000001)) 
```

#left-truncated imputation
```{r}
labres<- exposure_subset |>
  mutate(lod.flag = if_else(analysis_result < analysis_llod, 1, 0),
         test.result = analysis_result,
         log2.test.result = log2(analysis_result),
         test.name = analysis_analyte_name,
         test.time = "one")

LODmeansd.all <- labres %>% 
  group_by(test.name, test.time) %>% 
  summarise(meanlog2 = mean(log2.test.result, na.rm= TRUE), sdlog2 = sd(log2.test.result, na.rm=TRUE)) %>% 
  mutate(test.name = (paste0(test.name, ".t", test.time)))

dat_bbep <- labres |> 
  filter(test.name == "bis(butoxyethyl) phosphate")

dat_behp <- labres |> 
  filter(test.name == "bis(2-ethylhexyl) phosphate")

set.seed(1010)
dat_imp_bbep <- impute(dat_bbep, LODmeansd.all, "bis(butoxyethyl) phosphate.tone", "result")
dat_imp_behp <- impute(dat_behp, LODmeansd.all, "bis(2-ethylhexyl) phosphate.tone", "result")
```

#join imputed datasets
```{r}
dat_imp_behp <- dat_imp_behp |> mutate(analyte = "bis(2-ethylhexyl) phosphate")
dat_imp_bbep <- dat_imp_bbep |> mutate(analyte = "bis(butoxyethyl) phosphate")

imputed_dat<- bind_rows(dat_imp_behp, dat_imp_bbep)
write_csv(imputed_dat, "~/Documents/ECHO/imputed_left_trunc.csv")
```

#prep for viz
```{r}
viz_dat<- imputed_dat |>
  mutate(log2_result= log2(result.flag))

# Plotting the density plot with histogram faceted by analysis_analyte_name
viz_dat |>
  ggplot(aes(x = log2_result)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 1, fill = "skyblue", color = "skyblue", alpha = 0.5) +
  geom_density(color = "royalblue", linewidth=0.5) +
  labs(x = "Biomarker concentration on log2 scale", y = "Density",
       caption = " ") +
  ggtitle("") +
  #scale_y_continuous(trans = "exp")+
  #annotation_logticks(sides="lr")+
  scale_x_continuous(breaks = seq(-12, 5, by = 2), labels = seq(-12, 5, by = 2))+
  facet_wrap(~ analyte) +
  theme_minimal()+
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.caption = element_text(hjust = 0)  # Left-align the caption
  )+
  geom_vline(xintercept = -5.64, color="black")

ggsave("~/Documents/ECHO/exposure_viz.tiff",
       width=8, height= 4, dpi=300)
```

# Calculate the standard deviation of result.flg grouped by analyte
```{r}
std_dev_result <- imputed_dat |>
  group_by(analyte) |>
  summarise(std_dev_result.flg = sd(log2(result.flag), na.rm = TRUE))

dat_imp_bbep %>%
  filter(log2(result.flag) > -5.64 & log2(result.flag) < -3.86) %>%
  count()

dat_imp_bbep %>%
  filter(log2(result.flag) < -5.64 & log2(result.flag) > -7.42) %>%
  count()
```




```{r}
# Define x-axis intercepts for each facet panel
x_intercepts <- data.frame(
  analyte = c("bis(2-ethylhexyl) phosphate", "bis(butoxyethyl) phosphate"),
  x_intercept = c(-5.643856, -5.643856)
)

# Plotting the density plot with histogram faceted by analysis_analyte_name
p <- viz_dat |>
  ggplot(aes(x = log2_result)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 0.1, fill = "skyblue", color = "skyblue", alpha = 0.5) +
  geom_density(color = "royalblue", linewidth=1) +
  labs(x = "Biomarker concentration on log2 scale", y = "Density",
       caption = " ") +
  ggtitle("") +
  #scale_y_continuous(trans = "exp")+
  #annotation_logticks(sides="lr")+
  scale_x_continuous(breaks = seq(-12, 5, by = 3), labels = seq(-12, 5, by = 3))+
  facet_wrap(~ analyte) +
  theme_minimal()+
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.caption = element_text(hjust = 0)  # Left-align the caption
  )

# Add vertical lines for each facet panel
p + geom_vline(data = x_intercepts, aes(xintercept = x_intercept), linetype = "dashed", color = "red", linewidth = 0.7, alpha = 0.8)

ggsave("~/Documents/ECHO/exposure_viz.tiff",
       width=10, height= 8, dpi=300)
```

```{r}
x <- exp_dat |>
  mutate(LOD_binary = ifelse(analysis_result > analysis_llod, 1, 0)) |>
  mutate(analysis_analyte_name = fct_recode(analysis_analyte_name,
                                            "bis(butoxyethyl) phosphate" = "bis(butoxyethyl) pohosphate",
                                            "bis(2-ethylhexyl) phosphate" = "bis(2-ethylhexyl) phoshate")) |>
  filter(analysis_analyte_name %in% c("bis(1-chloro-2-propyl) phosphate")) |>
  filter(!is.na(analysis_result)) |>
  filter(!is.na(analysis_llod)) |>
  select(c(10,13,16,17,19))
```


